Validation of Visual Statistical Inference, Applied to Linear Models

### 2013-05-19 paper_jasa.tex
large p small n paper reference updated

%\documentclass{article}
\documentclass[12pt]{article}
\usepackage{times}
%\usepackage{natbib}
%\usepackage{multicol}
\RequirePackage{natbib}
\usepackage{amsmath, amssymb, fullpage, amsthm, array, algorithm2e,graphicx,asa}
%\usepackage[dvips]{graphics}

%\usepackage{hyperref} % for hyper reference

@@

\usepackage{color}
\newcommand{\blue}[1]{{\color{blue} #1}} %MM
\newcommand{\red}[1]{{\color{red} #1}} 
\newcommand{\green}[1]{{\color{green} #1}} %DC

\definecolor{orange}{rgb}{1,0.5,0}
\newcommand{\hh}[1]{{\color{orange} #1}} %HH

% \usepackage{pifont} % this package is used to print check mark \checkmark
% \linespread{1.6} % factor 1.6 = double space

@@

%\topmargin -0.10in   % when making pdf
%\textheight 9.15in  % when making pdf

@@

\begin{figure}[htp]
%\begin{figurehere}
   \centering
%       \includegraphics[width=0.95\textwidth]{lineup_category.pdf}
       \includegraphics[width=0.95\textwidth]{plot_turk1_300_10_12_1.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the largest shift or location difference between the boxplots? (The position of the actual data plot is provided in Section \ref{sec:category}.)}
       \label{fig:test_category}
\end{figure}
%\end{figurehere}

@@

The paper is organized as follows. Section \ref{sec:visual_test} defines terms as used in visual inference, and describes how to estimate the important quantities from experimental data. The effect of the lineup size and number of observers on the power of the test is discussed in Section \ref{sec:size}. Section \ref{sec:regression} focuses on the application of  visual inference to linear models.  Section \ref{sec:simulation} describes  three  user studies based on simulation experiments conducted to compare the power of the lineup protocol with the equivalent conventional test and  Section \ref{sec:results} presents an analysis of the resulting data. %The last section describes the next steps in this research.

@@

\begin{table}[htp]
\caption{Possible $p$-values for different numbers of observers, $K$, for fixed size $m = 20$ lineups.}
\begin{center}
\scalebox{.9}{
\begin{tabular}{|rrr|rrr|rrr|rrr|rrr|}\hline
$K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$  & $x$ & $p$-value & $K$  & $x$ & $p$-value\\ \hline %\cline{1-3}\cline{4-6}\cline{7-9}\cline{10-12}\cline{17-19}
1 &  1 & 0.0500 & 2 &  1 & 0.0975 & 3 & 1 & 0.1426 & 4 & 1 & 0.1855 & 5 & 1 & 0.2262 \\%\cline{1-3}
&&& 2 &  2 & 0.0025 & 3 & 2 & 0.0073 & 4 & 2 & 0.0140 & 5 & 2 & 0.0226 \\
&&& &&& 3 & 3 & 0.0001 & 4 & 3 & 0.0005 & 5 & 3 & 0.0012 \\%\cline{1-6}
&&&  &    &           &   &        && 4 & 4 & $< 0.0001$ & 5 & 4 & $< 0.0001$ \\\hline
\end{tabular}
}
\end{center}
\label{pvalue}
\end{table}

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=3in]{images/powerplot.pdf} 
   \caption{Probability that the data plot has the smallest probability in a lineup of size $m$. With increasing $p$-value the probability drops -- when it reaches $1/m$ a horizontal line is drawn to emphasize insufficient sensitivity of the test due to the lineup size. }
   \label{fig:pval_power}
\end{figure}

@@

In this setting, there are many established graphics that are used to evaluate and diagnose the fit of a regression model (e.g. \citealt{cook:99}). Table \ref{tbl:stat_multiple} lists several common hypotheses related to the regression setting, and commonly used statistical plots that might be used as corresponding visual test statistics.
%These plots should be familiar, because they are all in common use. 
For example, to examine the effect of variable $X_j$ on $Y$, we would plot residuals obtained from fitting the model without $X_j$ against $X_j$ or for a single covariate we may plot $Y$ against $X_j$ (cases 1-4 in Table~\ref{tbl:stat_multiple}). To assess whether the  assumptions of linearity is appropriate we would draw a plot of residuals against fitted values (case 5 in Table~\ref{tbl:stat_multiple}). For the purpose of comparing visual against conventional inference, we focus on cases 2 and 3, with a continuous and categorical explanatory variable, respectively.

%Note that some charts can be associated with several different testing situations. In our paper, we pick the examples for situation 2 and 3 as candidates.

@@

\begin{equation}\label{power_exp} 
%   Power(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
 p(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test for different $K$ (number of observers), obtained using $p(\beta)$ from Equation \ref{power_exp}. Notice that the expected power of the visual test exceeds the power of the conventional test as $K$  increases, and when $\beta$ gets larger.   Conversely, visual power is below conventional power for parameter values close to the null hypothesis. This is even more pronounced for large number of observers. At the same time, the point of intersection between visual and conventional power approaches the value of the null hypothesis as the number of observers approaches infinity, leading to an asymptotically perfect power curve of zero in the null hypothesis and one for any alternative value. We observe this dichotomy of visual power  in power estimates based on the data collected from user experiments, too. It features prominently in  figure \ref{fig:power_loess_effect}.

@@

%\green{Something probably needs to be said about subjects who do multiple tests, what checks are done.} \blue{We do have some data where we notice multiple responses for a lineup by a subject. This happens when subject accidentally use browser's ``back'' button and same data is saved twice. We made sure in the data cleaning process that we do not include more than one response on a lineup by a subject. For this we only keep the very first response.}

@@

Each subject viewed at least 10 lineups with the option to evaluate more.
%and we allowed them to provide more evaluations if they wanted. 
Depending on the parameter combinations we group the lineups in different difficulty levels  as easy, medium, hard and mixed (actual numbers are given in the supplementary material). For each difficulty level a specific number of lineups was randomly picked  for evaluation. This number is chosen so that total number of evaluations for each lineup for that group exceed the threshold $n_{\gamma}$. To satisfy this plan we needed to recruit at least 300 subjects.

\begin{table}[hbtp]
\caption{Combination of parameter values, $\beta_2$,  $n$ and $\sigma$, used for the simulation experiments.} % title name of the table
\centering
\scalebox{.9}{
\begin{tabular}{c c c c c}
% creating 10 columns
\hline\hline

@@

\begin{figure}[p]
%\begin{figurehere}
   \centering
%       \includegraphics[width=0.95\textwidth]{lineup_continuous.pdf}
       \includegraphics[width=0.95\textwidth]{plot_turk2_100_25_5_3.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope? (The position of the actual data plot is provided in Section \ref{sec:continuous}.)}
       \label{fig:test_continuous}
\end{figure}

@@

\[
  Y_i = \left\{
  \begin{array}{l l}
    \alpha+\beta X_i + \epsilon_i  & \quad  X_i \sim N(0,1) \quad  i =1,...,n\\
    \lambda+ \eta_i & \quad X_i \sim N(\mu,1/3) \quad  i=1, ...,n_c\\
  \end{array} \right.
\]
where $\epsilon_i \stackrel{iid}\sim N(0,\sigma)$, $\eta_i \stackrel{iid}\sim N(0,\sigma/3)$ and $\mu = -1.75$. $n_c$ is the size of the contaminated data. For the experiment we consider $n=100$ and $n_c=15$ producing actual data with $115$ points. Further, $\alpha=0$, $\lambda=10$, and $\sigma$ is chosen to be approximately 3.5, so that error standard deviation across both groups of the data is $5$. %Figure \ref{fig:cont_dat} shows an example of contaminated data, generated by this model for $\beta=5$.  
A linear model (Equation \ref{multi} with $p=1$  and intercept $\beta_0=0$) is fit to the contaminated data. This experiment  follows the test in Table \ref{tbl:stat_multiple}, case 2. The actual data plot shows a scatterplot of the residuals vs $X_1$, and the null  plots are scatterplots of null data generated by plotting simulated residuals from $N(0, {\hat{\sigma}}^2$)  against $X_1$.

@@

\begin{figure}[p]
%\begin{figurehere}
   \centering
%       \includegraphics[width=0.95\textwidth]{lineup_contaminated.pdf}
       \includegraphics[width=0.95\textwidth]{plot_turk3_100_40_5_5.png}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous but the inclusion of some contamination with the data spoils the normality assumption of error structure. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope? (The position of the actual data plot is provided in Section \ref{sec:contamination}.)}
       \label{fig:test_contaminated}
\end{figure}

@@

% In this process of collecting data some participants did not provide demographic information as we see in Table \ref{tbl:summary} that male and female participants do not add up to total participants. It is not clear why they did not provide the information. Now the concern arises whether we should keep their responses in the study. It is also possible that participants were not serious enough to provide feedback. 
%For each of the experiments we had a test plot (easy plot) for which they should have a correct evaluation without much effort.  This could be a criteria to determine their attentiveness in the study. For this we consider six different data cleaning criteria as shown below;

Amazon Mechanical Turk workers are paid for their efforts, not substantially, but on the scale of the minimum wage in the USA. Some workers will try to maximize their earnings for minimum effort, which can affect the results from the data. For example, some workers may simply randomly pick a plot, without actively examining the plots in the lineup. For the purpose of identifying these participants and cleaning the data, we use one of the very easy lineups that everybody was exposed to as a {\it reference lineup} and  take action based on a subject's answer to this reference:
if the subject failed to identify the actual data plot on the reference lineup, we remove all of this subject's data from the analysis. If the answer on the reference lineup is correct, we remove the answer for this lineup from the analysis, but keep all of the remaining answers. Table~\ref{tbl:summary} tabulates the number of subjects, genders and lineups evaluated after applying the data screening procedure.  A summary of all the data along with cleaned data are added to the supplementary material.

@@

This paper has demonstrated that statistical graphics can be used in statistical inference and validates the lineup protocol proposed by \citet{buja:2009}. Specific terminology was defined, and methods for obtaining the $p$-value and estimating the power of visual tests were introduced.  In order to calculate the theoretical power, it was assumed that observers will select the plot having the strongest signal in the lineup, and the experimental data suggests that for most observers this assumption holds.  Results from visual inference in the controlled setting of the simulation study are comparable to those obtained by conventional inference. %, in the setting where %conventional testing can be done. 
Visual inference is intended to provide valid tests where 
no conventional test exists, and our experiments in a controlled scenario suggest that it will perform as expected in the intended applications. The power of a visual test increases with the number of observers, which interestingly, leads to a result that the theoretical power of visual test can be better than that of conventional tests.

@@

Visual inference has been successfully used in two practical applications: to evaluate the power of competing graphical designs \citep{heike:2012}, and to detecting signal presence in large $p$, small $n$ data \citep{niladri:2012}. It is hoped that the lineup protocol will prove to be  valuable in data mining applications, and exploratory analyses, where there are no existing gauges of statistical significance.

%\paragraph{Acknowledgement:}
%This work was funded in part by National Science Foundation grant DMS 1007697.

@@

%\bibliographystyle{plain}
%\bibliographystyle{plainnat}
%\bibliographystyle{ieeetr}
\bibliographystyle{asa}
\bibliography{references}

%\end{multicols}

\end{document}  %End of document.

>>>>

\documentclass[12pt]{article}
\usepackage{times}

\RequirePackage{natbib}
\usepackage{amsmath, amssymb, amsthm, array, graphicx,asa}

@@

@@

@@

\begin{figure}[htp]
   \centering
       \includegraphics[width=0.95\textwidth]{plot_turk1_300_10_12_1.pdf}
       \caption{Lineup plot ($m=20$) using side-by-side boxplots for testing $H_0: \beta_k=0$. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the largest shift or location difference between the boxplots? (The position of the actual data plot is provided in Section \ref{sec:category}.)}
       \label{fig:test_category}
\end{figure}

@@

The paper is organized as follows. Section \ref{sec:visual_test} defines terms as used in visual inference, and describes how to estimate the important quantities from experimental data. The effect of the lineup size and number of observers on the power of the test is discussed in Section \ref{sec:size}. Section \ref{sec:regression} focuses on the application of  visual inference to linear models.  Section \ref{sec:simulation} describes  three  user studies based on simulation experiments conducted to compare the power of the lineup protocol with the equivalent conventional test and  Section \ref{sec:results} presents an analysis of the resulting data.

@@

\begin{table}[htp]
\caption{Possible $p$-values for different numbers of observers, $K$, for fixed size $m = 20$ lineups.}
\begin{center}
\scalebox{.9}{
\begin{tabular}{|rrr|rrr|rrr|rrr|rrr|}\hline
$K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$  & $x$ & $p$-value & $K$  & $x$ & $p$-value\\ \hline 
1 &  1 & 0.0500 & 2 &  1 & 0.0975 & 3 & 1 & 0.1426 & 4 & 1 & 0.1855 & 5 & 1 & 0.2262 \\%\cline{1-3}
&&& 2 &  2 & 0.0025 & 3 & 2 & 0.0073 & 4 & 2 & 0.0140 & 5 & 2 & 0.0226 \\
&&& &&& 3 & 3 & 0.0001 & 4 & 3 & 0.0005 & 5 & 3 & 0.0012 \\%\cline{1-6}
&&&  &    &           &   &        && 4 & 4 & $< 0.0001$ & 5 & 4 & $< 0.0001$ \\\hline
\end{tabular}
}
\end{center}
\label{pvalue}
\end{table}

@@

\begin{figure}[htbp]  
   \centering
   \includegraphics[width=3in]{images/powerplot.pdf} 
   \caption{Probability that the data plot has the smallest probability in a lineup of size $m$. With increasing $p$-value the probability drops -- when it reaches $1/m$ a horizontal line is drawn to emphasize insufficient sensitivity of the test due to the lineup size. }
   \label{fig:pval_power}
\end{figure}

@@

In this setting, there are many established graphics that are used to evaluate and diagnose the fit of a regression model (e.g. \citealt{cook:99}). Table \ref{tbl:stat_multiple} lists several common hypotheses related to the regression setting, and commonly used statistical plots that might be used as corresponding visual test statistics. For example, to examine the effect of variable $X_j$ on $Y$, we would plot residuals obtained from fitting the model without $X_j$ against $X_j$ or for a single covariate we may plot $Y$ against $X_j$ (cases 1-4 in Table~\ref{tbl:stat_multiple}). To assess whether the  assumptions of linearity is appropriate we would draw a plot of residuals against fitted values (case 5 in Table~\ref{tbl:stat_multiple}). For the purpose of comparing visual against conventional inference, we focus on cases 2 and 3, with a continuous and categorical explanatory variable, respectively.

@@

\begin{equation}\label{power_exp} 
 p(\beta)=Pr(p_{D} < p_0)  \quad \text{for}  \quad \beta \ne 0
\end{equation}
Figure \ref{fig:power_expected} shows the power of the conventional test in comparison to the expected power of the visual test for different $K$ (number of observers), obtained using $p(\beta)$ from Equation \ref{power_exp}. Notice that the expected power of the visual test exceeds the power of the conventional test as $K$  increases, and when $\beta$ gets larger.   Conversely, visual power is below conventional power for parameter values close to the null hypothesis. This is even more pronounced for large number of observers. At the same time, the point of intersection between visual and conventional power approaches the value of the null hypothesis as the number of observers approaches infinity, leading to an asymptotically perfect power curve of zero in the null hypothesis and one for any alternative value. We observe this dichotomy of visual power  in power estimates based on the data collected from user experiments, too. It features prominently in  figure \ref{fig:power_loess_effect}.

@@

@@

Each subject viewed at least 10 lineups with the option to evaluate more. Depending on the parameter combinations we group the lineups in different difficulty levels  as easy, medium, hard and mixed (actual numbers are given in the supplementary material). For each difficulty level a specific number of lineups was randomly picked  for evaluation. This number is chosen so that total number of evaluations for each lineup for that group exceed the threshold $n_{\gamma}$. To satisfy this plan we needed to recruit at least 300 subjects.

\begin{table}[hbtp]
\caption{Combination of parameter values, $\beta_2$,  $n$ and $\sigma$, used for the simulation experiments.} 
\centering
\scalebox{.9}{
\begin{tabular}{c c c c c}
% creating 10 columns
\hline\hline

@@

\begin{figure}[p]
   \centering
       \includegraphics[width=0.95\textwidth]{plot_turk2_100_25_5_3.pdf}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope? (The position of the actual data plot is provided in Section \ref{sec:continuous}.)}
       \label{fig:test_continuous}
\end{figure}

@@

\[
  Y_i = \left\{
  \begin{array}{l l}
    \alpha+\beta X_i + \epsilon_i  & \quad  X_i \sim N(0,1) \quad  i =1,...,n\\
    \lambda+ \eta_i & \quad X_i \sim N(\mu,1/3) \quad  i=1, ...,n_c\\
  \end{array} \right.
\]
where $\epsilon_i \stackrel{iid}\sim N(0,\sigma)$, $\eta_i \stackrel{iid}\sim N(0,\sigma/3)$ and $\mu = -1.75$. $n_c$ is the size of the contaminated data. For the experiment we consider $n=100$ and $n_c=15$ producing actual data with $115$ points. Further, $\alpha=0$, $\lambda=10$, and $\sigma$ is chosen to be approximately 3.5, so that error standard deviation across both groups of the data is $5$. A linear model (Equation \ref{multi} with $p=1$  and intercept $\beta_0=0$) is fit to the contaminated data. This experiment  follows the test in Table \ref{tbl:stat_multiple}, case 2. The actual data plot shows a scatterplot of the residuals vs $X_1$, and the null  plots are scatterplots of null data generated by plotting simulated residuals from $N(0, {\hat{\sigma}}^2$)  against $X_1$.

@@

\begin{figure}[p]
   \centering
       \includegraphics[width=0.95\textwidth]{plot_turk3_100_40_5_5.png}
       \caption{Lineup plot ($m=20$) using scatter plots for testing $H_0: \beta_k=0$ where covariate $X_k$ is continuous but the inclusion of some contamination with the data spoils the normality assumption of error structure. One of these plots is the plot of the actual data, and the remaining are null plots, produced by simulating data from a null model that assumes $H_0$ is true. Which plot is the most different from the others, in the sense that there is the steepest slope? (The position of the actual data plot is provided in Section \ref{sec:contamination}.)}
       \label{fig:test_contaminated}
\end{figure}

@@

Amazon Mechanical Turk workers are paid for their efforts, not substantially, but on the scale of the minimum wage in the USA. Some workers will try to maximize their earnings for minimum effort, which can affect the results from the data. For example, some workers may simply randomly pick a plot, without actively examining the plots in the lineup. For the purpose of identifying these participants and cleaning the data, we use one of the very easy lineups that everybody was exposed to as a {\it reference lineup} and  take action based on a subject's answer to this reference:
if the subject failed to identify the actual data plot on the reference lineup, we remove all of this subject's data from the analysis. If the answer on the reference lineup is correct, we remove the answer for this lineup from the analysis, but keep all of the remaining answers. Table~\ref{tbl:summary} tabulates the number of subjects, genders and lineups evaluated after applying the data screening procedure.

@@

This paper has demonstrated that statistical graphics can be used in statistical inference and validates the lineup protocol proposed by \citet{buja:2009}. Specific terminology was defined, and methods for obtaining the $p$-value and estimating the power of visual tests were introduced.  In order to calculate the theoretical power, it was assumed that observers will select the plot having the strongest signal in the lineup, and the experimental data suggests that for most observers this assumption holds.  Results from visual inference in the controlled setting of the simulation study are comparable to those obtained by conventional inference. Visual inference is intended to provide valid tests where no conventional test exists, and our experiments in a controlled scenario suggest that it will perform as expected in the intended applications. The power of a visual test increases with the number of observers, which interestingly, leads to a result that the theoretical power of visual test can be better than that of conventional tests.

@@

Visual inference has been successfully used in two practical applications: to evaluate the power of competing graphical designs \citep{heike:2012}, and to detecting signal presence in large $p$, small $n$ data \citep{niladri:2011}. It is hoped that the lineup protocol will prove to be  valuable in data mining applications, and exploratory analyses, where there are no existing gauges of statistical significance.

@@

\bibliographystyle{asa}
\bibliography{references}

\end{document}
